---
title: "Placement 1 Listening Test Rasch Analysis"
output:
  pdf_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
#load packages
library(tidyverse)
library(psych)
library(eRm)
library(knitr)

#load data
d <- here::here("data/placement_1.csv") %>%
  read_csv()

items <- select(d, 5:39)

#CTT reliability
reliability <- psych::alpha(items)
reliability$total
```

The Placement 1 Listening Test was taken by `r nrow(items)` test-takers. The average raw score was `r round(mean(rowSums(items)), 2)` out of `r ncol(items)` total points (SD = `r round(sd(rowSums(items)), 2)`). The histogram below shows the distribution of raw total scores.

```{r histogram, echo=FALSE}
hist(rowSums(items))
```

# Rasch Analysis of the Placement 1 Listening Test

```{r rasch analysis, include=FALSE}
##run it
rasch <- RM(items)
rasch_person <- person.parameter(rasch)
rasch_person_ability <- rasch_person$thetapar
rasch_person_se.theta <- rasch_person$se.theta

#person separation
rasch_person_reliability <- SepRel(rasch_person)

#Item Stats
item_stats <- itemfit(rasch_person)
```

A dichotomous Rasch model was fit to the Placement 1 Listening Test. Table 1 summarizes the measurement of item easiness and person ability.


Table 1.

*Measurement Summary*

Facet  |  N            | Mean Measure                         | Mean S.E.
-------|---------------|--------------------------------------|----------------------------------------
Persons|`r nrow(items)`|`r round(mean(rasch_person_ability[[1]], na.rm = T), 2)`|`r round(mean(rasch_person_se.theta[[1]], na.rm = T),2)`
Items  |`r ncol(items)`|`r round(mean(rasch$betapar), 2)`     |`r round(mean(rasch$se.beta),2)`

The person separation reliability was `r rasch_person_reliability$sep.rel`. Figure 1 below presents the distribution of person ability and item easiness. Where persons and items are parallel, test-takers have a .50 probability of answering the item correctly. Items to the left are easier, and items to the right are more difficult.


```{r Figure1, echo=FALSE}
plotPImap(rasch)
```
*Figure 1.* Person and item map.

\pagebreak

Table 2 presents detailed item statistics for all 35 listening items. Ideally, both Infit and Outfit values should fall within the range of 0.7-1.3 (Wright & Linacre, 1994).

Table 2.

*Rasch Item Statistics*

```{r echo=FALSE}
kable(tibble(Item = names(item_stats$i.fit), Beta = round(rasch$betapar, 2), 
                 Infit = round(item_stats$i.infitMSQ, 2),
                 InfitZ = round(item_stats$i.infitZ, 2),
                 Outfit = round(item_stats$i.outfitMSQ, 2),
                 OutfitZ = round(item_stats$i.outfitZ, 2)))
```

