---
title: "Part 1: Starting with Data"
teaching: 40
exercises: 20
questions:
- "What is a data.frame?"
- "How can I read a complete csv file into R?"
- "How can I get basic summary information about my dataset?"
- "How can I change the way R treats strings in my dataset?"
- "Why would I want strings to be treated differently?"
- "How are dates represented in R and how can I change the format?"
objectives:
- "Describe what a data frame is."
- "Load external data from a .csv file into a data frame."
- "Summarize the contents of a data frame."
- "Describe the difference between a factor and a string."
- "Convert between strings and factors."
- "Reorder and rename factors."
- "Change how character strings are handled in a data frame."
- "Examine and change date formats."

keypoints:
- "Use read_csv to read tabular data in R."
- "Use factors to represent categorical data in R."
---


```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
source("../bin/download_data.R")
```

## Presentation of the test data

The dataset we will be working with for this lesson has scores for items
from a listening and reading placement test.



| column_name | description |
| ----------- | ----------- |
| ID | A unique ID for each test taker |
| names | Fake names of the test takers |
| origin | Fake areas of origin of the test takers |
| q*_list_mi | A main idea question on the listening test |
| q*_list_det | A detail question on the listening test |
| q*_list_inf | An inference question on the listening test |
| q*_list_prag | A pragmatics question on the listening test |
| q*_read_mi | A main idea question about on the reading test |
| q*_read_voc | A vocabulary question about on the reading test |
| q*_read_det | A detail question about on the reading test |
| q*_read_trog | A text organization/discourse question about on the reading test |
| q*_read_inf |  An inference question about on the reading test |
| q*_read_purp | A question about the author's purpose on the reading test |
| q*_an | An anchor question |



You are going load the data in R's memory using the function `read_csv()` from the **`readr`** package which is part of the **`tidyverse`**. So, before we can use the `read_csv()` function, we need to load the package. Also, if you recall, the missing data is encoded as "NULL" in the dataset. We'll tell it to the function, so R will automatically convert all the "NULL" entries in the dataset into `NA`.

```{r, eval=TRUE,  message=FALSE, purl=TRUE}
library(tidyverse)

test_results_1 <- read_csv("data/placement_1.csv")
```

This statement doesn't produce any output because, as you might recall,
assignments don't display anything. If we want to check that our data has been
loaded, we can see the contents of the data frame by typing its name: `test_results_1`.

```{r, results='hold', purl=FALSE}
test_results_1
## Try also
## View(test_results_1)
## head(test_results_1)
```

> ## Note
>
> `read_csv()` assumes that fields are delineated by commas, however, in several
> countries, the comma is used as a decimal separator and the semicolon (;) is
> used as a field delineator. If you want to read in this type of files in R,
> you can use the `read_csv2` function. It behaves exactly like `read_csv` but
> uses different parameters for the decimal and the field separators. If you are
> working with another format, they can be both specified by the user. Check out
> the help for `read_csv()` by typing `?read_csv` to learn more. There is also
> the `read_tsv()` for tab-separated data files, and `read_delim()` allows you
> to specify more details about the structure of your file.
{: .callout}

## What are data frames and tibbles?

Data frames are the _de facto_ data structure for tabular data in `R`, and what
we use for data processing, statistics, and plotting.

A data frame is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because columns are vectors,
each column must contain a single type of data (e.g., characters, integers,
factors). For example, here is a figure depicting a data frame comprising a
numeric, a character, and a logical vector.

![](../fig/data-frame.svg)

A data frame can be created by hand, but most commonly they are generated by the
functions `read_csv()` or `read_table()`; in other words, when importing
spreadsheets from your hard drive (or the web).

A tibble is an extension of `R` data frames used by the _tidyverse_. When the
data is read using `read_csv()`, it is stored in an object of class `tbl_df`,
`tbl`,  and `data.frame`. You can see the class of an object with

```{r results='hold', purl=FALSE}
class(test_results_1)
```

As a `tibble`, the type of data included in each column is listed in an
abbreviated fashion below the column names. For instance, here `ID` is a
column of doubles (abbreviated `<dbl>`), and `names` is a column of characters
(`<chr>`).

## Inspecting data frames

When calling a `tbl_df` object (like `test_results_1` here), there is already a lot of information about our data frame being displayed such as the number of rows, the number of columns, the names of the columns, and as we just saw the class of data stored in each column. However, there are functions to extract this information from data frames.  Here is a non-exhaustive list of some of these
functions. Let's try them out!

* Size:
* `dim(test_results_1)` - returns a vector with the number of rows in the first element,
and the number of columns as the second element (the **dim**ensions of
the object)
* `nrow(test_results_1)` - returns the number of rows
* `ncol(test_results_1)` - returns the number of columns

* Content:
* `head(test_results_1)` - shows the first 6 rows
* `tail(test_results_1)` - shows the last 6 rows

* Names:
* `names(test_results_1)` - returns the column names (synonym of `colnames()` for
`data.frame` objects)

* Summary:
* `str(test_results_1)` - structure of the object and information about the class,
length and content of each column
* `summary(test_results_1)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides data frames.



## Indexing and subsetting data frames

Our interviews data frame has rows and columns (it has 2 dimensions), if we want to
extract some specific data from it, we need to specify the "coordinates" we
want from it. Row numbers come first, followed by column numbers. However, note
that different ways of specifying these coordinates lead to results with
different classes.


```{r, purl=FALSE}
## first element in the first column of the data frame (as a vector)
test_results_1[1, 1]
## first element in the 6th column (as a vector)
test_results_1[1, 6]
## first column of the data frame (as a vector)
test_results_1[, 1]
## first column of the data frame (as a data.frame)
test_results_1[1]
## first three elements in the 7th column (as a vector)
test_results_1[1:3, 7]
## the 3rd row of the data frame (as a data.frame)
test_results_1[3, ]
## equivalent to head_test_results_1 <- head(test_results_1)
test_results_1 <- test_results_1[1:6, ]
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r, purl=FALSE}
test_results_1[, -1]          # The whole data frame, except the first column
test_results_1[-c(7:88), ]   # Equivalent to head(test_results_1)
```

Data frames can be subset by calling indices (as shown previously), but also by calling their column names directly:

```{r, eval = FALSE, purl=FALSE}
test_results_1["country"]       # Result is a data frame
test_results_1[, "country"]     # Result is a data frame
test_results_1[["country"]]     # Result is a vector
test_results_1$country          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct names of the columns.

> ## Exercise
>
> 1. Create a data frame (`test_results_1_50`) containing only the data in
>    row 50 of the `test_results_1` dataset.
>
> 2. Notice how `nrow()` gave you the number of rows in a data frame?
>
>      * Use that number to pull out just that last row in the data frame.
>      * Compare that with what you see as the last row using `tail()` to make
>        sure it's meeting expectations.
>      * Pull out that last row using `nrow()` instead of the row number.
>      * Create a new data frame (`test_results_1_last`) from that last row.
>
> 3. Use `nrow()` to extract the row that is in the middle of the data frame.
>    Store the content of this row in an object named `test_results_1_middle`.
>
> 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
>    `head(test_results_1)`, keeping just the first through 6th rows of the test results
>    dataset.
>
> > ## Solution
> >
> > ```{r}
> > ## 1.
> > test_results_1_50 <- test_results_1[50, ]
> > ## 2.
> > # Saving `n_rows` to improve readability and reduce duplication
> > n_rows <- nrow(test_results_1)
> > test_results_1_last <- test_results_1[n_rows, ]
> > ## 3.
> > test_results_1_middle <- test_results_1[n_rows / 2, ]
> > ## 4.
> > test_results_1_head <- test_results_1[-(7:n_rows), ]
> > ```
> {: .solution}
{: .challenge}


## Factors

R has a special data class, called factor, to deal with categorical data that
you may encounter when creating plots or doing statistical analyses. Factors are
very useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with
labels and they can be ordered or unordered. While factors look (and often
behave) like character vectors, they are actually treated as integer vectors by
R. So you need to be very careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
country <- factor(c("Peru", "Netherlands", "Chile", "Kenya"))
```

R will assign `1` to the level `"Chile"` and `4` to the level `"Peru"`
(because `c` comes before `p`, even though the first element in this vector is
`"Peru"`). You can see this by using the function `levels()` and you can find
the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(country)
nlevels(country)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `country` vector would be:

```{r, results=TRUE, purl=FALSE}
country # current order
country <- factor(country, levels = c("Peru", "Netherlands", "Chile", "Kenya"))
country # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2), but are more
informative than integers because factors are self describing: `"Chile"`,
`"Kenya"` is more descriptive than `1`, and `2`. Which one is "Chile"? You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels. It also makes renaming levels easier. Let's say we made a mistake and need to recode "Chile" to "Argentina".

```{r, results=TRUE, purl=FALSE}
levels(country)
levels(country)[3] <- "Argentina"
levels(country)
country
```



### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(country)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. The `as.numeric()`
function returns the index values of the factor, not its levels, so it will
result in an entirely new (and unwanted in this case) set of numbers.
One method to avoid this is to convert factors to characters, and then to numbers.
Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)                     # Wrong! And there is no warning...
as.numeric(as.character(year_fct))       # Works...
as.numeric(levels(year_fct))[year_fct]   # The recommended way.
```

Notice that in the recommended `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(year_fct)`
* We convert these levels to numeric values using `as.numeric(levels(year_fct))`
* We then access these numeric values using the underlying integers of the
vector `year_fct` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor level.
Let's extract the `country` column from our data frame, convert it into a
factor, and use it to look at the number of test takers from different
parts of the world:

```{r purl=TRUE, eval=FALSE}
## create a vector from the data frame column "country"
res_country <- test_results_1$country
## convert it into a factor
res_country <- as.factor(res_country)
## print the vector
res_country
## plot the vector
plot(res_country)
```


> ## Exercise
>
> * Rename the levels of the factor to have the first letter in uppercase:
>   "China","Morrocco", and "Russia".
>
> * Now that we have renamed the factor levels, can you
>   recreate the barplot such that the labels are in reverse alphabet order?
>
> > ## Solution
> >
> > ```{r factor-plot-exercise, purl=FALSE, eval=FALSE}
> > ## 1.
> > levels(res_country) <- c("China", "Morocco", "Russia")
> > ## or 2.
> > res_country <- factor(res_country, levels = c("Russia", "Morocco", "China"))
> > plot(res_country)
> ```
> {: .solution}
{: .challenge}


<!-- ## Formatting Dates -->

<!-- One of the most common issues that new (and experienced!) R users have is -->
<!-- converting date and time information into a variable that is appropriate and -->
<!-- usable during analyses. As a reminder from earlier in this lesson, the best -->
<!-- practice for dealing with date data is to ensure that each component of your -->
<!-- date is stored as a separate variable. In our dataset, we have a -->
<!-- column `admin_date` which contains information about the -->
<!-- year, month, and day that the test was administered. Let's -->
<!-- convert those dates into three separate columns. -->

<!-- ```{r, eval=FALSE, purl=FALSE} -->
<!-- str(test_results_1) -->
<!-- ``` -->
<!-- We are going to use the package **`lubridate`** (which belongs to the **`tidyverse`**; learn more [here](https://www.tidyverse.org/)) to work with dates. **`lubridate`** gets installed as part as the **`tidyverse`** installation. When you load  the **`tidyverse`** (`library(tidyverse)`), the core packages (the packages used in most data analyses) get loaded. **`lubridate`** however does not belong to the core tidyverse, so you have to load it explicitly with `library(lubridate)` -->

<!-- Start by loading the required package: -->

<!-- ```{r load-package, message=FALSE, purl=FALSE} -->
<!-- library(lubridate) -->
<!-- ``` -->

<!-- The lubridate function `ymd()` takes a vector representing year, month, and day, and converts it to a -->
<!-- `Date` vector. `Date` is a class of data recognized by R as being a date and can -->
<!-- be manipulated as such. The argument that the function requires is flexible, -->
<!-- but, as a best practice, is a character vector formatted as "YYYY-MM-DD". -->


<!-- Let's extract our `interview_date` column and inspect the structure: -->

<!-- ```{r, purl=FALSE} -->
<!-- dates <- test_results_1$admin_date -->
<!-- str(dates) -->
<!-- ``` -->

<!-- When we imported the data in R, `read_csv()` recognized that this column contained date information. We can now use the `day()`, `month()` and  `year()` functions to extract this information from the date, and create new columns in our data frame to store it: -->


<!-- ```{r, purl=FALSE} -->
<!-- test_results_1$day <- day(dates) -->
<!-- test_results_1$month <- month(dates) -->
<!-- test_results_1$year <- year(dates) -->
<!-- test_results_1 -->
<!-- ``` -->

<!-- Notice the three new columns at the end of our data frame. -->

{% include links.md %}
