---
title: "Part 1: Data visualisation with ggplot2"
teaching: 60
exercises: 35
questions:
  - "What are the components of a ggplot?"
  - "How do I create scatterplots, boxplots, and barplots?"
  - "How can I change the aesthetics (ex. colour, transparency) of my plot?"
  - "How can I create multiple plots at once?"
objectives:
   - "Produce scatter plots, boxplots, and time series plots using ggplot."
   - "Set universal plot settings."
   - "Describe what faceting is and apply faceting in ggplot."
   - "Modify the aesthetics of an existing ggplot plot (including axis labels and color)."
   - "Build complex and customized plots from data in a data frame."
keypoints:
   - "`ggplot2` is a flexible and useful tool for creating plots in R."
   - "The data set and coordinate system can be defined using the `ggplot` function."
   - "Additional layers, including geoms, are added using the `+` operator."
   - "Boxplots are useful for visualizing the distribution of a continuous variable."
   - "Barplot are useful for visualizing categorical data."
   - "Faceting allows you to generate multiple plots based on a categorical variable."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")
source("../bin/download_data.R")
```

We start by loading the required package. **`ggplot2`** is also included in the
**`tidyverse`** package.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

If not still in the workspace, load the data we saved in the previous lesson.


```{r load-data, purl=FALSE}
test_1_raw <- read_csv('data_output/test_1_raw_totals.csv') %>%
  mutate(country = as.factor(country)) # changes the country variable to a factor

# test_results_2 <- read_csv('data/placement_2.csv') %>%
#   mutate(country = as.factor(country)) # changes the country variable to a factor
```

## Plotting with **`ggplot2`**

**`ggplot2`** is a plotting package that makes it simple to create complex plots
from data stored in a data frame. It provides a programmatic interface for
specifying what variables to plot, how they are displayed, and general visual
properties. Therefore, we only need minimal changes if the underlying data
change or if we decide to change from a bar plot to a scatterplot. This helps in
creating publication quality plots with minimal amounts of adjustments and
tweaking.

**`ggplot2`** functions like data in the 'long' format, i.e., a column for every
dimension, and a row for every observation. Well-structured data will save you
lots of time when making figures with **`ggplot2`**

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

- use the `ggplot()` function and bind the plot to a specific data frame using
  the `data` argument

```{r ggplot-steps-1, eval=FALSE, purl=FALSE}
ggplot(data = test_1_raw)
```

- define a mapping (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

```{r ggplot-steps-2, eval=FALSE, purl=FALSE}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total))
```

- add 'geoms' – graphical representations of the data in the plot (points,
lines, bars). **`ggplot2`** offers many different geoms; we will use some
common ones today, including:

* `geom_point()` for scatter plots, dot plots, etc.
* `geom_boxplot()` for, well, boxplots!
* `geom_line()` for trend lines, time series, etc.

To add a geom to the plot use the `+` operator. Because we have two continuous variables,
let's use `geom_point()` first:

```{r first-ggplot, purl=FALSE}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
    geom_point()
```

The `+` in the **`ggplot2`** package is particularly useful because it allows
you to modify existing `ggplot` objects. This means you can easily set up plot
templates and conveniently explore different types of plots, so the above plot
can also be generated with code like this:

```{r first-ggplot-with-plus, eval=FALSE, purl=FALSE}
# Assign plot to a variable
sub_skill_scatter <- ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total))

# Draw the plot
sub_skill_scatter +
    geom_point()
```

> ## Notes
>
> - Anything you put in the `ggplot()` function can be seen by any geom layers
>   that you add (i.e., these are universal plot settings). This includes the x-
>   and y-axis mapping you set up in `aes()`.
> - You can also specify mappings for a given geom independently of the mapping
>   defined globally in the `ggplot()` function.
> - The `+` sign used to add new layers must be placed at the end of the line
>   containing the *previous* layer. If, instead, the `+` sign is added at the
>   beginning of the line containing the new layer, **`ggplot2`** will not add
>   the new layer and will return an error message.
{: .callout}

```{r ggplot-with-plus-position, eval=FALSE, purl=FALSE}
## This is the correct syntax for adding layers
sub_skill_scatter +
    geom_point()

## This will not add the new layer and will return an error message
sub_skill_scatter
+ geom_point()
```

## Building your plots iteratively

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:
	
```{r create-ggplot-object, purl=FALSE}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
  geom_point()
```

Then, we start modifying this plot to extract more information from it. For
instance, we can add transparency (`alpha`) to avoid overplotting:

```{r adding-transparency, purl=FALSE}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
    geom_point(alpha = 0.5)
```

That only helped a little bit with the overplotting problem. We can also
introduce a little bit of randomness into the position of our points
using the `geom_jitter()` function.

```{r adding-jitter, purl=FALSE}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
    geom_jitter(alpha = 0.5)
```


We can also add colors for all the points:

```{r adding-colors, purl=FALSE}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
  geom_jitter(alpha = 0.5, color = "blue")
```

Or to color dots by country of origin in the plot differently, you could use a vector as an input to the argument **`color`**.  Because we are now mapping features of the data to a color, instead of setting one color for all points, the color now needs to be set inside a call to the **`aes`** function. **`ggplot2`** will provide a different color corresponding to different values in the vector. We set the value of **`alpha`** outside of the **`aes`** function call because we are using the same value for all points. Here is an example where we color by **`country`**:


```{r color-by-species, purl=FALSE}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
    geom_jitter(aes(color = country), alpha = 0.5)
```

There is a positive relationship between the raw scores on the reading and listening tests. There does
not seem to be a difference in this relationship by country of origin.

> ## Exercise
>
> Use what you just learned to create a scatter plot of `read_an_raw` by `list_an_raw`
> with the `country` showing in different colors. Is this a good
> way to show this type of data?
>
> > ## Solution
> >
> > ```{r scatter-challenge, answer=TRUE, purl=FALSE}
> > ggplot(data = test_1_raw, aes(x = read_an_raw, y = list_an_raw)) +
> >    geom_jitter(aes(color = country), alpha = 0.5)
> > ```
> {: .solution}
{: .challenge}


## Boxplot

We can use boxplots to visualize the distribution of total raw test scores 
across countries of origin:

```{r boxplot, purl=FALSE}
ggplot(data = test_1_raw, aes(x = country, y = raw_total)) +
    geom_boxplot()
```

By adding points to a boxplot, we can have a better idea of the number of
measurements and of their distribution:

```{r boxplot-with-points, purl=FALSE}
ggplot(data = test_1_raw, aes(x = country, y = raw_total)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.5, color = "tomato", width = .20)
```


Notice how the boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot in front of the points such that it's not
hidden?

> ## Exercise
>
> Boxplots are useful summaries, but hide the *shape* of the distribution. For
> example, if the distribution is bimodal, we would not see it in a
> boxplot. An alternative to the boxplot is the violin plot, where the shape
> (of the density of points) is drawn.
>
> - Replace the box plot with a violin plot; see `geom_violin()`.
> 
> > ## Solution
> >
> > ```{r violin-plot}
> > ggplot(data = test_1_raw, aes(x = country, y = raw_total)) +
> >   geom_violin(alpha = 0) +
> >   geom_jitter(alpha = 0.5, color = "tomato")
> > ```
> {: .solution}
{: .challenge}

## Barplots

Barplots are also useful for visualizing categorical data. By default,
`geom_bar` accepts a variable for x, and plots the number of instances each
value of x (in this case, wall type) appears in the dataset.

```{r barplot-1}
ggplot(data = test_1_raw, aes(x = country)) +
    geom_bar()
```

Let's say that, hypothetically, people who recieved a raw score of greater than a 50 are admitted into 
the university, or some kind of program. We can include `mutate`, `case_when`, `%>%`, and `ggplot` to create a new
variable called `admitted`. Then we can plot it.

```{r, barplot-2}
admit_bar <- test_1_raw %>%
  mutate(admitted = case_when(raw_total > 50 ~ 'yes',
                              raw_total <= 50 ~ 'no')) %>%
  ggplot(., aes(x = country, fill = admitted)) +
  geom_bar(position = 'dodge') # places the bars next to each other

```


> ## Exercise
>
> Pretend that in order to be admitted, test takers need a reading subscore
> 25 or greater. Create a bar plot showing the counts of respondents from
> each country of origin who recieved a raw reading score greater or equal to
> than 25. 
>
> > ## Solution
> >
> > ```{r barplot-memb-assoc}
> > admit_bar_read <- test_1_raw %>%
> >  mutate(admitted_read = case_when(read_raw_total >= 25 ~ 'yes',
> >                              read_raw_total < 25 ~ 'no')) %>%
> >  ggplot(., aes(x = country, fill = admitted_read)) +
> >  geom_bar(position = 'dodge') # places the bars next to each other
> > ```
> >
> {: .solution}
{: .challenge}


## Adding Labels and Titles

By default, the axes labels on a plot are determined by the name of the variable
being plotted. However, **`ggplot2`** offers lots of customization options,
like specifying the axes labels, and adding a title to the plot with 
relatively few lines of code. We will add more informative x and y axis
labels to our plot of proportion of house type by village and also add
a title.

```{r scatter-labeled}
ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
    geom_jitter(aes(color = country), alpha = 0.5) +
    xlab("Total Raw Reading Scores") +
    ylab("Total Raw Listening Scores") +
    ggtitle("Listening <> Reading Subscores") +
    scale_color_viridis_d(labels = c('China', 'Morocco', 'Russia'), # add caps to key
                          name = 'Country') # add caps to legend title

```

## Faceting

Rather than creating a single plot with side-by-side bars or points for each
country of origin, we may want to create multiple plot, where each plot shows the
data for a single country of origin. This would be especially useful if we had
a large number of villages that we had sampled, as a large number of
side-by-side bars will become more difficult to read.

**`ggplot2`** has a special technique called *faceting* that allows the user to split one
plot into multiple plots based on a factor included in the dataset. We
will use it to split our scatterplot of of listening and reading scores
by country of origin so that each country of origin has it's own panel in a multi-panel plot:

```{r barplot-faceting, fig.height=6, fig.width=12}
title_case <- as_labeller(c('china' = 'China',
                            'morocco' = 'Morocco',
                            'russia' = 'Russia'))

ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
    geom_jitter(alpha = 0.5) +
    xlab("Total Raw Reading Scores") +
    ylab("Total Raw Listening Scores") +
    ggtitle("Listening <> Reading Subscores") +
    facet_wrap(~ country, nrow = 1, labeller = title_case)
```

Click the "Zoom" button in your RStudio plots pane to view a larger
version of this plot.

Usually plots with white background look more readable when printed.  We can set
the background to white using the function `theme_bw()`. Additionally, you can remove
the grid:

```{r barplot-theme-bw, purl=FALSE, fig.height=6, fig.width=12}
title_case <- as_labeller(c('china' = 'China',
                            'morocco' = 'Morocco',
                            'russia' = 'Russia'))

ggplot(data = test_1_raw, aes(x = read_raw_total, y = list_raw_total)) +
    geom_jitter(alpha = 0.5) +
    xlab("Total Raw Reading Scores") +
    ylab("Total Raw Listening Scores") +
    ggtitle("Listening <> Reading Subscores") +
    facet_wrap(~ country, nrow = 1, labeller = title_case) +
    theme_bw(base_size = 18) + # increase base font size to 18
    theme(panel.grid = element_blank())
```

What if we wanted to see histograms of the total scores on the test by country of origin:

```{r percent-items-data}
total_score_hist <- test_1_raw %>%
  ggplot(., aes(x = raw_total)) +
  geom_histogram() +
  facet_wrap(~ country, labeller = title_case) +
  ylab('Counts') +
  xlab('Total Raw Test Score') +
  xlim(0, 70) + # set the limits of the x-axis at 0 and 70 (min and max possible)
  ggtitle('Total Raw Test Score Distribution by Country')
```


## **`ggplot2`** themes

In addition to `theme_bw()`, which changes the plot background to white,
**`ggplot2`** comes with several other themes which can be useful to quickly
change the look of your visualization. The complete list of themes is available
at <http://docs.ggplot2.org/current/ggtheme.html>. `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html)
package provides a wide variety of options (including an Excel 2003 theme). The
[**`ggplot2`** extensions website](https://www.ggplot2-exts.org) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.

> ## Exercise
>
> Experiment with at least two different themes. Build the previous plot
> using each of those themes. Which do you like best?
{: .challenge}

## Customization

Take a look at the [**`ggplot2`** cheat
sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf),
and think of ways you could improve the plot.

> ## Exercise
>
> With all of this information in hand, please take another five minutes to
> either improve one of the plots generated in this exercise or create a
> beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)
> for inspiration. Here are some ideas:
>
> * See if you can make the bars white with black outline.
> * Try using a different color palette (see
>   http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/).
{: .challenge}

After creating your plot, you can save it to a file in your favorite format. The Export tab in the **Plot** pane in RStudio will save your plots at low resolution, which will not be accepted by many journals and will not scale well for posters.

Instead, use the `ggsave()` function, which allows you easily change the dimension and resolution of your plot by adjusting the appropriate arguments (`width`, `height` and `dpi`).

Make sure you have the `fig_output/` folder in your working directory.

```{r ggsave-example, eval=FALSE, purl=FALSE}
ggsave("fig_output/name_of_file.png", total_score_hist, width = 6, height = 4, units = 'in')
```

Note: The parameters `width` and `height` also determine the font size in the saved plot.

{% include links.md %}
